---
title: "Is alcohol consumption related to gender and/or GPA?"
author: "Put your name here!"
date: "put the date here"
output: html_document
---

```{r include=FALSE,echo=FALSE}
require(tigerstats)
```

# Introduction


This study is to see if alcohol consumption in the ucdavis1 data set differs by sex and/or GPA. 
1. Do men or women drink more?
2. Is alchol consumption related to GPA?
3. Do 'C' students show an alcohol consumption difference based on gender
4. What about 'A' students?

The ucdavis1 data set is included in the tigerstats package. It is a data frame with 173 observations on 12 variables. It was taken from the text book used at UC Irvine, Mind on Statistics, 4th edition, Uts and Heckard. The variables and descriptions are as follows.
Description

Results of a survey of students at UC-Davis.

* **Sex** a factor with levels Female Male

* **TV** Number of hours spent watching TV per week

* **computer** number of hours spent on computer per week

* **Sleep** hours of sleep per night

* **Seat** a factor with levels Back Front Middle Where do you prefer to sit in class, when you have a choice?

* **alcohol** number of alcoholic drinks consumed per week

* **Height** height in inches

* **momheight** height of mother, in inches

* **dadheight** height of father, in inches

* **exercise** number of hours of exercise per week

* **GPA** grade point average

* **class** a factor with levels LibArts NonLib Student Category: liberal arts or not

Doing a variable analysis shows that the varibles of interest will be

**alcohol**  a numeric variable
**Sex** a factor variable 
and **GPA** a numeric variable that we will need to do further work with in order to answer question pertaining to **A** and **C** students


<In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?>

# Methods

The questions are all somewhat related. Questions 1,3, and 4 are asking the same question, the only difference being the members of the groups the question references: 

* all students
* A students or
* C students

The question deals with a categorical variable (**Sex**) modeling differences in a continuous variable (**alcohol**). 

Question 2 is relating two continuous variables **alcohol** and **GPA** which require regression/correlation techniques.

So our approach will be to deal with all four question in each block of this report.

Boxplots and t-tests for questions 1,3, and 4
scatterplot and regression for question 2.

We see that problems 1, 3, and 4 are basically the same question being asked of different groups. The methods we use for these problems involve explaining a numerical variable (**alcohol**) with the categorical variable (**Sex**).  We will use boxplots and or violin plots as our graphical tool, and t-tests as our inferential methods.
To deal with questions about A and C students we will use this definition
* **A**  GPA greater than 3.5
* **C**  GPA between 2 and 2.75
To answer question 2 we note that both GPA and alcohol are continuous variables. Our graphical tools scatterplots and regression line fits.  So our approach will be to address all four questions in each of our major sections and summarize all results in the conclusion section

<Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.>

#Results

First we create two smaller data frames. One called **C.Students** and the other **A.Students** by limiting membership to just those students with qualifying GPA scores.

```{r}
require(tigerstats)
all.Students <- na.omit(ucdavis1)
cvals=(all.Students$GPA>=2)&(all.Students$GPA<=2.75)
avals=(all.Students$GPA>3.5)
C.Students <-all.Students[cvals,]
A.Students <-all.Students[avals,]
Males.all<-all.Students[all.Students$Sex=="Male",]
Females.all<-all.Students[all.Students$Sex=="Female",]
C.Females<-C.Students[C.Students$Sex=="Female",]
C.Males<-C.Students[C.Students$Sex=="Male",]
A.Males<-A.Students[A.Students$Sex=="Male",]
A.Females<-A.Students[A.Students$Sex=="Female",]
```
### Overall summary of Students 

There are a total of 173 students, 64 of them we are labeling **C.Students** and 29 as **A.Students** 94 **Females** and 79 **Males** 

=======
The questions are all somewhat related. Questions 1,3, and 4 are asking the same question, the only difference being the members of the groups the question references: 

* all students
* A students or
* C students

The question deals with a categorical variable (**Sex**) modeling differences in a continuous variable (**alcohol**). 

Question 2 is relating two continuous variables **alcohol** and **GPA** which require regression/correlation techniques.

So our approach will be to deal with all four question in each block of this report.
Boxplots and violin plots as graphics and t-tests for inferential for questions 1,3, and 4

scatterplot and regression for question 2.



#Results


<Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.>

## Descriptive Results

### Graphical Descriptive Results


#### Question 1

We first look at a boxplot with an overlayed violin plot of all students.

```{r}
bwplot(alcohol ~ Sex,
	data = all.Students,
	panel = function(..., box.ratio) {
		panel.bwplot(..., fill = NULL, box.ratio = 0.1)
		panel.violin(..., col = "burlywood", border = "grey60", alpha = 0.5, varwidth = FALSE)},
	main = "Alcohol Consumption All Students",
	ylab = "Alcoholic drinks per week")
```

Looking at our graphical display it appears that male students are drinking more tha females.  Moreover the variation of the amount imbibed seems larger for the males as well.  Look at the size of the box in the plot of the male data. Its vertical dimention is twice as large as the box for the females.  This can be interpreted as an approximate comparison of the variance of the two groups of drinkers with that of the males being twice that of females.

#### Question 2

To compare alcohol and GPA both of which are continuous variables we resort to a scatterplot.

```{r}
xyplot(alcohol ~ GPA,
	data = all.Students,
	type = c("p","r"),
	main = "Alcohol consumption by GPA")

```

The regression line seems to show a negative association between GPA and amount of drinking. Higher GPA is correlated with lower alcohol consumption. We will next display the same sort of plot, but also taking sex into account conditionally.

```{r}
xyplot(alcohol ~ GPA,
	data = all.Students,
	groups = Sex,
	auto.key = list(
		space = "top",
		title = "Sex",
		cex.title = 1,
		columns = 1),
	type = c("p","r"),
	main = "Alcohol consumption vs GPA conditioned by Sex")
```

This plot is more revealing!  It appears that GPA and drinking are negatively correlated fore male students, but not so much for females.

#### Question 3
We look at a boxplot with an overlayed violin plot of C students.

```{r}
bwplot(alcohol ~ Sex,
	data = C.Students,
	panel = function(..., box.ratio) {
		panel.bwplot(..., fill = NULL, box.ratio = 0.1)
		panel.violin(..., col = "burlywood", border = "grey60", alpha = 0.5, varwidth = FALSE)},
	main = "Alcohol Consumption C.Students",
	ylab = "Alcoholic drinks per week")
```

The plot shows the inner quartile range of the **male** students (shown by the height of the rectangle in the plot) to be about four times as great as that of the **females** thus raising doubt that male and female drinking behavior is similiar

#### Question 4

We now look at a boxplot with an overlayed violin plot of A students.

```{r}
bwplot(alcohol ~ Sex,
	data = A.Students,
	panel = function(..., box.ratio) {
		panel.bwplot(..., fill = NULL, box.ratio = 0.1)
		panel.violin(..., col = "burlywood", border = "grey60", alpha = 0.5, varwidth = FALSE)},
	main = "Alcohol Consumption A Students",
	ylab = "Alcoholic drinks per week")
```

The plot shows the inner quartile range of the **male** students (shown by the height of the rectangle in the plot) to be comperable to that of the **females** 


<Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question>

### Numerical Descriptive Results

#### Question 1

We will display summary statistics for men and women separately

```{r}
favstats(alcohol,data=Males.all,na.rm=TRUE)
favstats(alcohol,data=Females.all,na.rm=TRUE)
```

These results show a numeric difference in both center and spread of alcohol consumption and spread for Males and females as well. The Male average number of drinks is three times that of Females.  The variation measured by the respective standard deviations is also different by a factor of three with standard deviation of male alcohol consumption being over 10, while that of females around 3.3

#### Question 2

#### Question 3

```{r}
favstats(alcohol,data=C.Males,na.rm=TRUE)
favstats(alcohol,data=C.Females,na.rm=TRUE)
```
These results show a large numeric difference in both center and spread of alcohol consumption and spread for Males and females who are C students. The Male average number of drinks is four times that of Females.  The variation measured by the respective standard deviations is also different by a factor of four with standard deviation of male alcohol consumption being over 14, while that of females around 3.5

#### Question 4

```{r}
favstats(alcohol,data=A.Males,na.rm=TRUE)
favstats(alcohol,data=A.Females,na.rm=TRUE)
```


<Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question>

## Inferential Results

<State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question>


# Discussion and Conclusion

<Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.>
